---
title: "lol-modeling"
author: "Vladislav Fedorov"
date: "2026-07-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# eda

```{r}
library(readr)
library(purrr)
library(tidyr)
library(dplyr)

lol_data <- list.files("data/raw", full.names = TRUE) |>
  map_dfr(read_csv, guess_max = Inf)

dim(lol_data)
```


```{r}
names(lol_data)

one_game <- lol_data |> filter(gameid == first(gameid))
one_game |> select(gameid, side, position, playername, teamname, champion, kills, deaths, assists, result) |> print(n = 12)
```

```{r}
# confirm they're truly identical, not just similar
identical(lol_data$dragons, lol_data$`dragons (type unknown)`)

# find out which specific file(s) actually contain the oddly-named column
walk(list.files("data/raw", full.names = TRUE), function(f) {
  cols <- names(read_csv(f, n_max = 0, show_col_types = FALSE))
  if ("dragons (type unknown)" %in% cols) cat(f, "\n")
})
```

```{r}
# do dragons and the "unknown type" column ever both have values in the same row,
# or are they mutually exclusive (suggesting one is a fallback for the other)?
lol_data |>
  summarise(
    both_na      = sum(is.na(dragons) & is.na(`dragons (type unknown)`)),
    dragons_only = sum(!is.na(dragons) & is.na(`dragons (type unknown)`)),
    unknown_only = sum(is.na(dragons) & !is.na(`dragons (type unknown)`)),
    both_present = sum(!is.na(dragons) & !is.na(`dragons (type unknown)`))
  )

# does it line up with data completeness specifically?
lol_data |>
  count(datacompleteness, has_unknown_dragons = !is.na(`dragons (type unknown)`))
```

```{r}
lol_data <- lol_data |>
  select(-`dragons (type unknown)`)
```

```{r}
lol_data |>
  group_by(datacompleteness) |>
  summarise(pct_na_golddiffat15 = mean(is.na(golddiffat15)), n = n())
```

```{r}
lol_data |>
  group_by(datacompleteness) |>
  summarise(
    pct_na_result = mean(is.na(result)),
    pct_na_gamelength = mean(is.na(gamelength)),
    pct_na_kills = mean(is.na(kills)),
    pct_na_dragons = mean(is.na(dragons)),
    n = n()
  )
```
```{r}
10/12
```

```{r}
lol_data |>
  filter(position == "team") |>
  group_by(datacompleteness) |>
  summarise(pct_na_dragons = mean(is.na(dragons)), n = n())
```

```{r}
lol_data |>
  filter(datacompleteness == "complete") |>
  group_by(position) |>
  summarise(pct_na_golddiffat15 = mean(is.na(golddiffat15)), n = n())
```

```{r}
one_game_teams <- lol_data |>
  filter(position == "team", gameid == first(gameid)) |>
  select(gameid, side, result, goldat15, opp_goldat15, golddiffat15)

print(one_game_teams)
```

```{r}
complete_gameid <- lol_data |>
  filter(position == "team", datacompleteness == "complete") |>
  slice(1) |>
  pull(gameid)

lol_data |>
  filter(position == "team", gameid == complete_gameid) |>
  select(gameid, side, result, goldat15, opp_goldat15, golddiffat15)
```

# pre modeling features

```{r}
lol_data |>
  filter(position == "team", gameid == complete_gameid) |>
  select(gameid, side, result, firstblood, firstdragon, firsttower, firstherald, firstbaron)


```

```{r}
win_prediction_data <- lol_data |>
  filter(position == "team", datacompleteness == "complete") |>
  select(gameid, side, teamname, league, year, patch, result,
         golddiffat15, xpdiffat15, csdiffat15,
         firstblood, firstdragon, firsttower, firstherald)

win_prediction_data |>
  summarise(across(c(golddiffat15, xpdiffat15, csdiffat15, firstblood, firstdragon, firsttower, firstherald),
                    ~ mean(is.na(.))))
```

# modeling

```{r}
library(xgboost)
```

```{r}
model_2024 <- glm(result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
                   data = win_prediction_data |> filter(year == 2024) |> na.omit(), family = binomial)
model_2025 <- glm(result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
                   data = win_prediction_data |> filter(year == 2025) |> na.omit(), family = binomial)

cbind(coef(model_2024), coef(model_2025))
```

Noticeable difference between effect size between 2024/5. However, it is defensible that 2024 and 2025 are still valuable for prediction of results in 2026.

```{r}
train_data <- win_prediction_data |> filter(year %in% c(2024, 2025)) |> na.omit()
test_data  <- win_prediction_data |> filter(year == 2026) |> na.omit()

nrow(train_data)
nrow(test_data)
```

```{r}
feature_cols <- c("golddiffat15", "xpdiffat15", "csdiffat15", "firstblood", "firstdragon", "firsttower", "firstherald")

train_matrix <- train_data |> select(all_of(feature_cols)) |> as.matrix()
test_matrix  <- test_data  |> select(all_of(feature_cols)) |> as.matrix()

dtrain <- xgb.DMatrix(data = train_matrix, label = train_data$result)
dtest  <- xgb.DMatrix(data = test_matrix, label = test_data$result)
```

```{r}
model <- glm(
  result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
  data = train_data, family = binomial
)

summary(model)
```

```{r}
library(pROC)

test_data <- test_data |>
  mutate(pred_prob = predict(model, newdata = test_data, type = "response"),
         pred_class = ifelse(pred_prob > 0.5, 1, 0))

accuracy <- mean(test_data$pred_class == test_data$result)
roc_obj <- roc(test_data$result, test_data$pred_prob)

accuracy
auc(roc_obj)
```

```{r}
mean(train_data$result)  
```

# xgboost

```{r}

params <- xgb.params(
  objective = "binary:logistic"
)

set.seed(1)
cv <- xgb.cv(
  params = params,
  data = dtrain, nrounds = 500, nfold = 5,
  metrics = list("auc"),
  early_stopping_rounds = 20,
  maximize = TRUE,
  verbose = 0
)

best_nrounds <- cv$early_stop$best_iteration
best_nrounds
```

```{r}
xgb_model <- xgb.train(
  params = params,
  data = dtrain, nrounds = best_nrounds
)

test_data <- test_data |>
  mutate(pred_prob_xgb = predict(xgb_model, dtest))

roc_xgb <- roc(test_data$result, test_data$pred_prob_xgb)
auc(roc_xgb)
```

```{r}
xgb.importance(model = xgb_model)
```

Using early-game state (gold, XP, and CS differentials at 15 minutes, plus first-objective indicators) from professional League of Legends match data, a logistic regression model trained on 2024–2025 data achieved an AUC of 0.842 on a held-out 2026 test set. A gradient boosting comparison model achieved a nearly identical AUC of 0.840, indicating the relationship between early-game state and match outcome is close to linear.

Coefficient stability was checked across training years to assess sensitivity to patch changes. Gold and experience differentials remained stable predictors, while first-objective indicators shifted in importance between years; first blood specifically lost statistical significance by 2025 and showed no predictive value in either model. Gold differential at 15 minutes was the dominant predictor, accounting for approximately 85 percent of the gradient boosting model's total contribution.

# weekly gold difference at 15 minutes 

```{r}
library(lubridate)

weekly_series <- lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = floor_date(date, "week")) |>
  group_by(week) |>
  summarise(
    mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE),
    n_games = n()
  ) |>
  arrange(week)

range(weekly_series$week)
nrow(weekly_series)

summary(weekly_series$n_games)
weekly_series |> filter(n_games < 10) |> print(n = Inf)
```

```{r}
library(tsibble)
library(zoo)

weekly_series <- lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  group_by(week) |>
  summarise(
    mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE),
    n_games = n()
  ) |>
  arrange(week)

range(weekly_series$week)
nrow(weekly_series)
summary(weekly_series$n_games)
```
```{r}
weekly_ts <- weekly_series |>
  mutate(mean_abs_golddiff15 = if_else(n_games < 10, NA_real_, mean_abs_golddiff15)) |>
  as_tsibble(index = week) |>
  fill_gaps()

sum(is.na(weekly_ts$n_games))

weekly_ts <- weekly_ts |>
  mutate(mean_abs_golddiff15 = na.approx(mean_abs_golddiff15, na.rm = FALSE))

sum(is.na(weekly_ts$mean_abs_golddiff15))
```
```{r}
weekly_ts |> filter(is.na(mean_abs_golddiff15))
```

```{r}
weekly_ts <- weekly_ts |> filter(!is.na(mean_abs_golddiff15))

sum(is.na(weekly_ts$mean_abs_golddiff15))
nrow(weekly_ts)
```

```{r}
library(ggplot2)
weekly_ts |>
  ggplot(aes(x = week, y = mean_abs_golddiff15)) +
  geom_line() +
  geom_point(data = weekly_ts |> filter(n_games < 10 | is.na(n_games)), color = "red") +
  labs(title = "Weekly mean |gold diff at 15 min|", x = NULL, y = "Gold difference")
```

```{r}
weekly_ts |> as_tibble() |> arrange(desc(mean_abs_golddiff15)) |> slice(1:3)
weekly_ts |> as_tibble() |> arrange(mean_abs_golddiff15) |> slice(1:3)
```

```{r}
low_week_start <- as.Date(min(weekly_ts$week[weekly_ts$mean_abs_golddiff15 == min(weekly_ts$mean_abs_golddiff15)]))

lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue",
         date >= low_week_start, date < low_week_start + days(7)) |>
  count(league, patch, sort = TRUE)
```


```{r}
high_week_start <- as.Date(yearweek("2024 W41"))
lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue",
         date >= high_week_start, date < high_week_start + days(7)) |>
  count(league, patch, sort = TRUE)
```

```{r}
# top cluster, checked together as one connected block
lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  filter(week %in% yearweek(c("2024 W40", "2024 W41", "2024 W42"))) |>
  group_by(week, league) |>
  summarise(mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE), n = n(), .groups = "drop") |>
  arrange(week, desc(n))

# bottom cluster, checked separately since they're not consecutive
lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  filter(week %in% yearweek(c("2025 W27", "2025 W42", "2025 W47"))) |>
  group_by(week, league) |>
  summarise(mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE), n = n(), .groups = "drop") |>
  arrange(week, desc(n))
```

```{r}
lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  count(league, sort = TRUE) |>
  print(n = 20)
```

# modeling 2

```{r}
library(fable)
library(feasts)
```

```{r}
weekly_ts |>
  ACF(mean_abs_golddiff15, lag_max = 20) |>
  autoplot()

weekly_ts |>
  gg_tsdisplay(mean_abs_golddiff15, plot_type = "partial")
```
```{r}
weekly_ts |> features(mean_abs_golddiff15, unitroot_kpss)
```
KPSS test did not reject stationarity 

```{r}
fit <- weekly_ts |>
  model(
    arima_auto = ARIMA(mean_abs_golddiff15),
    arima_manual = ARIMA(mean_abs_golddiff15 ~ pdq(1, 0, 0)),
    ets = ETS(mean_abs_golddiff15)
  )

fit |> select(arima_auto) |> report()
```

```{r}
fit |> select(arima_manual) |> report()
```

```{r}
weekly_cv <- weekly_ts |>
  stretch_tsibble(.init = 78, .step = 4)

cv_forecasts <- weekly_cv |>
  model(
    arima_auto   = ARIMA(mean_abs_golddiff15),
    arima_manual = ARIMA(mean_abs_golddiff15 ~ pdq(1, 0, 0)),
    ets          = ETS(mean_abs_golddiff15)
  ) |>
  forecast(h = 4)

cv_forecasts |> accuracy(weekly_ts)
```


# xgboost on arima(3, 0, 1) model

```{r}
weekly_features <- weekly_ts |>
  as_tibble() |>
  arrange(week) |>
  mutate(
    lag1 = lag(mean_abs_golddiff15, 1),
    lag2 = lag(mean_abs_golddiff15, 2),
    lag3 = lag(mean_abs_golddiff15, 3),
    lag4 = lag(mean_abs_golddiff15, 4)
  )

# the first 4 rows will have NA lags by construction, since there's no history to lag from yet
sum(is.na(weekly_features$lag4))

weekly_features <- weekly_features |> filter(!is.na(lag4))
nrow(weekly_features)
```

```{r}
library(xgboost)

origins <- seq(78, nrow(weekly_features) - 4, by = 4)

fold_results <- map_dfr(origins, function(origin) {
  train_fold <- weekly_features |> slice(1:origin)
  test_fold  <- weekly_features |> slice((origin + 1):(origin + 4))

  dtrain_fold <- xgb.DMatrix(
    data = as.matrix(train_fold |> select(lag1, lag2, lag3, lag4)),
    label = train_fold$mean_abs_golddiff15
  )
  dtest_fold <- xgb.DMatrix(
    data = as.matrix(test_fold |> select(lag1, lag2, lag3, lag4))
  )

  xgb_fold <- xgb.train(
    params = xgb.params(objective = "reg:squarederror"),
    data = dtrain_fold, nrounds = 50, verbose = 0
  )

  tibble(
    week = test_fold$week,
    actual = test_fold$mean_abs_golddiff15,
    pred = predict(xgb_fold, dtest_fold)
  )
})

# same error metrics as the classical models, computed the same way
fold_results |>
  summarise(
    RMSE = sqrt(mean((actual - pred)^2)),
    MAE  = mean(abs(actual - pred))
  )
```

```{r}
arima_errors <- cv_forecasts |>
  filter(.model == "arima_auto") |>
  as_tibble() |>
  left_join(as_tibble(weekly_ts) |> select(week, actual = mean_abs_golddiff15), by = "week") |>
  mutate(arima_error = abs(actual - .mean)) |>
  select(week, arima_error)

xgb_errors <- fold_results |>
  mutate(xgb_error = abs(actual - pred)) |>
  select(week, xgb_error)

error_compare <- arima_errors |>
  inner_join(xgb_errors, by = "week") |>
  mutate(xgb_advantage = arima_error - xgb_error) |>
  arrange(desc(xgb_advantage))

error_compare |> print(n = 15)
```

Weekly mean early-game gold differential (|golddiffat15|) was constructed as a proxy for how decisive professional games are by the 15-minute mark, then forecast using classical and machine learning approaches with rolling-origin backtesting. Extreme weeks were investigated directly to identify their underlying causes. Unusually high weeks were driven by smaller regional leagues, while unusually low weeks corresponded to international events (MSI, Worlds), where closely matched competition produces less decisive early games. This league-composition effect is documented as a known limitation of the pooled series.

Model order was identified via ACF/PACF inspection and confirmed against automatic ARIMA selection. A manually specified AR(1) model had comparable in-sample AIC to the selected ARIMA(3,0,1) but performed meaningfully worse on out-of-sample backtesting (RMSE 327 vs. 312), illustrating that in-sample fit criteria do not reliably predict forecasting accuracy. ARIMA(3,0,1) and an ETS model performed comparably as the strongest classical approaches.

A gradient boosting model using lagged features was subsequently compared against both using matched backtesting folds, achieving a notably lower RMSE (274.7 vs. 311.7), with MAE remaining comparable. This advantage was traced to specific weeks surrounding major international tournaments and season transitions, periods during which the series behaves least like a smooth continuous process, consistent with a nonlinear model more effectively capturing sharp deviations from trend that a linear autoregressive structure underestimates.

# leagues comparison

```{r}
lol_data |>
  filter(position == "team") |>
  count(league, sort = TRUE) |>
  print(n = 30)
```

```{r}
lcs_teams  <- lol_data |> filter(league == "LCS", position == "team") |> distinct(teamname) |> pull()
ltan_teams <- lol_data |> filter(league == "LTA N", position == "team") |> distinct(teamname) |> pull()
intersect(lcs_teams, ltan_teams)

lla_cblol_teams <- lol_data |> filter(league %in% c("LLA", "CBLOL"), position == "team") |> distinct(teamname) |> pull()
ltas_teams      <- lol_data |> filter(league == "LTA S", position == "team") |> distinct(teamname) |> pull()
intersect(lla_cblol_teams, ltas_teams)
```

```{r}
region_map <- tibble(
  league = c("LPL", "LCK", "LEC", "LCS", "LTA N", "LLA", "CBLOL", "LTA S"),
  region = c("China", "Korea", "Europe", "North America", "North America",
             "Latin America", "Latin America", "Latin America")
)

team_home_region <- lol_data |>
  filter(position == "team", league %in% region_map$league) |>
  left_join(region_map, by = "league") |>
  count(teamname, region, sort = TRUE) |>
  group_by(teamname) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup()

team_home_region |> count(region, sort = TRUE)
nrow(team_home_region)
```

```{r}
game_results <- lol_data |>
  filter(position == "team") |>
  select(gameid, date, side, teamname, result) |>
  pivot_wider(names_from = side, values_from = c(teamname, result)) |>
  mutate(winner = if_else(result_Blue == 1, teamname_Blue, teamname_Red)) |>
  select(gameid, date, team_blue = teamname_Blue, team_red = teamname_Red, winner) |>
  arrange(date)

nrow(game_results)
sum(is.na(game_results$date))
```
# elo model


```{r}
run_elo <- function(K) {
  teams <- unique(c(game_results$team_blue, game_results$team_red))
  ratings <- setNames(rep(1500, length(teams)), teams)
  n_games <- nrow(game_results)
  e_blue_vec <- numeric(n_games)
  actual_blue_win <- numeric(n_games)

  for (i in seq_len(n_games)) {
    blue <- game_results$team_blue[i]
    red  <- game_results$team_red[i]
    r_blue <- ratings[[blue]]
    r_red  <- ratings[[red]]

    e_blue <- 1 / (1 + 10^((r_red - r_blue) / 400))
    s_blue <- if (game_results$winner[i] == blue) 1 else 0

    e_blue_vec[i] <- e_blue
    actual_blue_win[i] <- s_blue

    ratings[[blue]] <- r_blue + K * (s_blue - e_blue)
    ratings[[red]]  <- r_red + K * ((1 - s_blue) - (1 - e_blue))
  }

  mean((e_blue_vec - actual_blue_win)^2)
}

results_by_K <- map_dfr(c(8, 16, 24, 32, 40, 48, 56), function(k) {
  tibble(K = k, brier_score = run_elo(k))
})

results_by_K
```

```{r}
run_elo_full <- function(K) {
  teams <- unique(c(game_results$team_blue, game_results$team_red))
  ratings <- setNames(rep(1500, length(teams)), teams)
  n_games <- nrow(game_results)

  for (i in seq_len(n_games)) {
    blue <- game_results$team_blue[i]
    red  <- game_results$team_red[i]
    r_blue <- ratings[[blue]]
    r_red  <- ratings[[red]]
    e_blue <- 1 / (1 + 10^((r_red - r_blue) / 400))
    s_blue <- if (game_results$winner[i] == blue) 1 else 0
    ratings[[blue]] <- r_blue + K * (s_blue - e_blue)
    ratings[[red]]  <- r_red + K * ((1 - s_blue) - (1 - e_blue))
  }

  tibble(teamname = names(ratings), rating = unname(ratings))
}

ratings_24 <- run_elo_full(24) |> rename(rating_24 = rating)
ratings_48 <- run_elo_full(48) |> rename(rating_48 = rating)

compare_ratings <- ratings_24 |> left_join(ratings_48, by = "teamname")

cor(compare_ratings$rating_24, compare_ratings$rating_48, method = "spearman")

compare_ratings |> arrange(desc(rating_24)) |> slice(1:10) |> select(teamname, rating_24)
compare_ratings |> arrange(desc(rating_48)) |> slice(1:10) |> select(teamname, rating_48)
```

```{r}
game_results |>
  filter(team_blue == "T1" | team_red == "T1") |>
  mutate(t1_won = (team_blue == "T1" & winner == "T1") | (team_red == "T1" & winner == "T1")) |>
  arrange(date) |>
  mutate(game_num = row_number()) |>
  group_by(period = ntile(game_num, 4)) |>
  summarise(win_rate = mean(t1_won), n = n())
```

keep K = 24

```{r}
region_ratings <- team_home_region |>
  left_join(ratings_24, by = "teamname")

region_ratings |>
  group_by(region) |>
  summarise(
    mean_rating = mean(rating_24),
    median_rating = median(rating_24),
    max_rating = max(rating_24),
    n_teams = n()
  ) |>
  arrange(desc(mean_rating))

region_ratings |>
  ggplot(aes(x = reorder(region, rating_24, median), y = rating_24)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "Team Elo ratings by region", x = NULL, y = "Elo rating")
```

```{r}
# a cleaner proxy for "elite tier strength" than a mean dragged around by weak teams
region_ratings |>
  group_by(region) |>
  slice_max(rating_24, n = 3) |>
  summarise(top3_mean = mean(rating_24), .groups = "drop") |>
  arrange(desc(top3_mean))

# how much uncertainty actually surrounds each region's mean, given how uneven n_teams is
set.seed(1)
bootstrap_region_means <- map_dfr(1:2000, function(i) {
  region_ratings |>
    group_by(region) |>
    slice_sample(prop = 1, replace = TRUE) |>
    summarise(mean_rating = mean(rating_24), .groups = "drop")
})

bootstrap_region_means |>
  group_by(region) |>
  summarise(lower_95 = quantile(mean_rating, 0.025), upper_95 = quantile(mean_rating, 0.975)) |>
  arrange(desc(lower_95))
```

Team strength was estimated using a sequentially updated Elo rating system spanning all professional matches, including domestic league play and international events (MSI, Worlds), which serve as the primary bridge for comparing teams across regions that otherwise rarely compete against one another. Prior to model construction, a structural change in the Americas competitive scene was identified and confirmed through team-roster overlap: LCS, LLA, and CBLOL were consolidated into a unified LTA system mid-dataset, requiring these leagues to be treated as continuous regions.

The K-factor governing rating sensitivity was evaluated empirically across a range of values using one-step-ahead prediction accuracy (Brier score) instead of a default assumption. Results indicated the top competitive tier is stable regardless of K-factor choice, while mid-table rankings show meaningful sensitivity to how heavily recent results are weighted. K=24, the conventional value, was retained given its comparable predictive performance and standard interpretability.

Regional comparison was conducted using bootstrapped confidence intervals instead of point estimates alone. Raw mean ratings suggested a clear regional hierarchy led by Korea; however, confidence intervals for all five regions overlapped substantially, indicating this apparent ranking is not statistically well-supported given the limited number of teams per region. A top-3 average provided a more defensible measure of elite-tier strength, indicating Korea's advantage over other regions is concentrated in genuine depth at the top. China's high maximum rating, by contrast, reflects a single standout team instead of broad regional strength.

# patch hypothesis testing

```{r}
raw_patches <- map_dfr(list.files("data/raw", full.names = TRUE), function(f) {
  read_csv(f, col_types = cols_only(gameid = col_character(), patch = col_character()))
})

raw_patches |> distinct(patch) |> arrange(patch) |> print(n = Inf)
```

```{r}
raw_patches |>
  distinct(gameid, .keep_all = TRUE) |>
  summarise(n_missing = sum(is.na(patch)), n_total = n(), pct_missing = mean(is.na(patch)))
```


```{r}
patch_herald_tower <- lol_data |>
  filter(position == "team", side == "Blue", !is.na(patch)) |>
  group_by(patch) |>
  summarise(
    win_rate_given_herald = mean(result[firstherald == 1], na.rm = TRUE),
    win_rate_given_tower  = mean(result[firsttower == 1], na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(patch)

# same sparse-sample caution from Part 2 — a handful of games isn't enough to trust a win rate
summary(patch_herald_tower$n_games)
patch_herald_tower |> filter(n_games > 30) |> print(n = Inf)

patch_herald_tower |>
  pivot_longer(cols = c(win_rate_given_herald, win_rate_given_tower), names_to = "objective", values_to = "win_rate") |>
  ggplot(aes(x = patch, y = win_rate, color = objective, group = objective)) +
  geom_line() +
  geom_point() +
  labs(title = "Win rate given first herald / first tower, by patch", x = "Patch", y = "Win rate")
```

```{r}
patch_herald_tower_clean <- patch_herald_tower |>
  filter(n_games >= 30) |>
  arrange(patch) |>
  mutate(patch_factor = factor(patch, levels = unique(patch), ordered = TRUE))

patch_herald_tower_clean |>
  pivot_longer(cols = c(win_rate_given_herald, win_rate_given_tower), names_to = "objective", values_to = "win_rate") |>
  ggplot(aes(x = patch_factor, y = win_rate, color = objective, group = objective)) +
  geom_line() +
  geom_point() +
  labs(title = "Win rate given first herald / first tower, by patch (n \u2265 30 games)", x = "Patch", y = "Win rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
win_prediction_patched <- win_prediction_data |>
  mutate(season = floor(patch))

season_coefs <- map_dfr(c(14, 15, 16), function(s) {
  m <- glm(
    result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
    data = win_prediction_patched |> filter(season == s), family = binomial
  )
  tibble(season = s, term = names(coef(m)), estimate = coef(m))
})

season_coefs |> filter(term %in% c("firsttower", "firstherald")) |> arrange(term, season)
```

```{r}
boundary_data <- win_prediction_patched |>
  filter(season %in% c(14, 15)) |>
  mutate(period = factor(season, levels = c(14, 15), labels = c("before", "after")))

m_null <- glm(
  result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald + period,
  data = boundary_data, family = binomial
)

m_tower_int  <- update(m_null, . ~ . + firsttower:period)
m_herald_int <- update(m_null, . ~ . + firstherald:period)

tower_test  <- anova(m_null, m_tower_int, test = "Chisq")
herald_test <- anova(m_null, m_herald_int, test = "Chisq")

tower_p  <- tower_test$`Pr(>Chi)`[2]
herald_p <- herald_test$`Pr(>Chi)`[2]

p.adjust(c(firsttower = tower_p, firstherald = herald_p), method = "fdr")
```

This analysis investigated an anomaly identified in Part 1: the predictive weight of first herald and first tower on match outcome nearly doubled between two training years, while first blood's effect collapsed to non-significance. Before conducting patch-level analysis, the integrity of the patch identifier was verified directly, confirming that zero-padded patch notation (e.g., 14.01, 14.10) avoids a numeric collision risk that would otherwise be introduced by floating-point storage of version numbers.

Season-level coefficients from Part 1's original model were recomputed using patch-derived season boundaries instead of calendar years. Both objectives showed a comparable increase in predictive weight between seasons 14 and 15, confirming the original finding was not an artifact of the calendar-year split. Season 16 revealed a divergence: first tower's elevated importance persisted, while first herald's effect partially reverted toward its original level, indicating two distinct underlying dynamics rather than a single shared cause.

The season 14-to-15 shift was formally tested using likelihood-ratio tests for objective-by-period interaction terms within the original logistic regression, the appropriate test given that the original finding concerned a controlled coefficient rather than a marginal win rate. Following FDR correction across both objectives, both interactions were highly significant (first tower: p = 2.47 × 10⁻¹³; first herald: p = 1.11 × 10⁻⁹), confirming the coefficient shifts identified in Part 1 reflect genuine changes in game dynamics rather than sampling variation.
