---
title: "League of Legends Esports Analysis"
author: "Vladislav Fedorov"
date: "2026-07-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Win Prediction

## Data loading and structure

```{r}
library(readr)
library(purrr)
library(tidyr)
library(dplyr)

# Oracle's Elixir professional match data, one CSV per year
lol_data <- list.files("data/raw", full.names = TRUE) |>
  map_dfr(read_csv, guess_max = Inf)

dim(lol_data)
```

```{r}
# each game is 12 rows: 2 team-level rows (position == "team") + 10 player-level rows
names(lol_data)

one_game <- lol_data |> filter(gameid == first(gameid))
one_game |> select(gameid, side, position, playername, teamname, champion, kills, deaths, assists, result) |> print(n = 12)
```

## Data quality checks

```{r}
# "dragons (type unknown)" turned out to be a duplicate of "dragons"
identical(lol_data$dragons, lol_data$`dragons (type unknown)`)

walk(list.files("data/raw", full.names = TRUE), function(f) {
  cols <- names(read_csv(f, n_max = 0, show_col_types = FALSE))
  if ("dragons (type unknown)" %in% cols) cat(f, "\n")
})
```

```{r}
# the duplicate only ever appears alongside a real "dragons" value, and only for
# partial-completeness games. The column is dropped in the next code chunk
lol_data |>
  summarise(
    both_na      = sum(is.na(dragons) & is.na(`dragons (type unknown)`)),
    dragons_only = sum(!is.na(dragons) & is.na(`dragons (type unknown)`)),
    unknown_only = sum(is.na(dragons) & !is.na(`dragons (type unknown)`)),
    both_present = sum(!is.na(dragons) & !is.na(`dragons (type unknown)`))
  )

lol_data |>
  count(datacompleteness, has_unknown_dragons = !is.na(`dragons (type unknown)`))
```

```{r}
lol_data <- lol_data |>
  select(-`dragons (type unknown)`)
```

```{r}
lol_data |>
  group_by(datacompleteness) |>
  summarise(pct_na_golddiffat15 = mean(is.na(golddiffat15)), n = n())
```

```{r}
# final results, per-player stats, and team objective counts are fully present in both
# completeness groups
lol_data |>
  group_by(datacompleteness) |>
  summarise(
    pct_na_result = mean(is.na(result)),
    pct_na_gamelength = mean(is.na(gamelength)),
    pct_na_kills = mean(is.na(kills)),
    pct_na_dragons = mean(is.na(dragons)),
    n = n()
  )
```

```{r}
# the ~83% NA rate above is a row-type issue, not a completeness issue: dragons is
# a team-level stat and is never populated on the 10 player rows per game
lol_data |>
  filter(position == "team") |>
  group_by(datacompleteness) |>
  summarise(pct_na_dragons = mean(is.na(dragons)), n = n())
```

```{r}
# golddiffat15 lives on team rows too, not just player rows
lol_data |>
  filter(datacompleteness == "complete") |>
  group_by(position) |>
  summarise(pct_na_golddiffat15 = mean(is.na(golddiffat15)), n = n())
```

```{r}
# the two team rows per game are exact mirror images of each other (own/opp stats,
# diff columns) 
complete_gameid <- lol_data |>
  filter(position == "team", datacompleteness == "complete") |>
  slice(1) |>
  pull(gameid)

lol_data |>
  filter(position == "team", gameid == complete_gameid) |>
  select(gameid, side, result, goldat15, opp_goldat15, golddiffat15)
```

## Feature preparation

```{r}
# first-objective indicators follow the same mirrored, team-perspective pattern
lol_data |>
  filter(position == "team", gameid == complete_gameid) |>
  select(gameid, side, result, firstblood, firstdragon, firsttower, firstherald, firstbaron)
```

```{r}
win_prediction_data <- lol_data |>
  filter(position == "team", datacompleteness == "complete") |>
  select(gameid, side, teamname, league, year, patch, result,
         golddiffat15, xpdiffat15, csdiffat15,
         firstblood, firstdragon, firsttower, firstherald)

# no leftover missingness in the feature set
win_prediction_data |>
  summarise(across(c(golddiffat15, xpdiffat15, csdiffat15, firstblood, firstdragon, firsttower, firstherald),
                    ~ mean(is.na(.))))
```

## Baseline model and year-over-year stability

```{r}
library(xgboost)
```

```{r}
# checked whether the early-game/win relationship held steady year to year
model_2024 <- glm(result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
                   data = win_prediction_data |> filter(year == 2024) |> na.omit(), family = binomial)
model_2025 <- glm(result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
                   data = win_prediction_data |> filter(year == 2025) |> na.omit(), family = binomial)

cbind(coef(model_2024), coef(model_2025))
```

Gold and XP differentials stayed stable across years, but firsttower/firstherald roughly
doubled in weight and firstblood's effect nearly vanished by 2025. Pooling 2024-2025 for
training is still defensible since the dominant predictor (golddiffat15) is the stable one;
the drift in the secondary indicators is investigated properly in Part 4.

```{r}
# chronological split: train on full seasons 2024-2025, test on 2026 
train_data <- win_prediction_data |> filter(year %in% c(2024, 2025)) |> na.omit()
test_data  <- win_prediction_data |> filter(year == 2026) |> na.omit()

nrow(train_data)
nrow(test_data)
```

```{r}
feature_cols <- c("golddiffat15", "xpdiffat15", "csdiffat15", "firstblood", "firstdragon", "firsttower", "firstherald")

train_matrix <- train_data |> select(all_of(feature_cols)) |> as.matrix()
test_matrix  <- test_data  |> select(all_of(feature_cols)) |> as.matrix()

dtrain <- xgb.DMatrix(data = train_matrix, label = train_data$result)
dtest  <- xgb.DMatrix(data = test_matrix, label = test_data$result)
```

```{r}
model <- glm(
  result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
  data = train_data, family = binomial
)

summary(model)
```

```{r}
library(pROC)

test_data <- test_data |>
  mutate(pred_prob = predict(model, newdata = test_data, type = "response"),
         pred_class = ifelse(pred_prob > 0.5, 1, 0))

accuracy <- mean(test_data$pred_class == test_data$result)
roc_obj <- roc(test_data$result, test_data$pred_prob)

accuracy
auc(roc_obj)
```

```{r}
# class balance check 
mean(train_data$result)
```

## Gradient boosting comparison

```{r}
params <- xgb.params(
  objective = "binary:logistic"
)

set.seed(1)
cv <- xgb.cv(
  params = params,
  data = dtrain, nrounds = 500, nfold = 5,
  metrics = list("auc"),
  early_stopping_rounds = 20,
  maximize = TRUE,
  verbose = 0
)

best_nrounds <- cv$early_stop$best_iteration
best_nrounds
```

```{r}
xgb_model <- xgb.train(
  params = params,
  data = dtrain, nrounds = best_nrounds
)

test_data <- test_data |>
  mutate(pred_prob_xgb = predict(xgb_model, dtest))

roc_xgb <- roc(test_data$result, test_data$pred_prob_xgb)
auc(roc_xgb)
```

```{r}
xgb.importance(model = xgb_model)
```

**Summary:** Using early-game state (gold, XP, and CS differentials at 15 minutes, plus
first-objective indicators) from professional League of Legends match data, a logistic
regression model trained on 2024-2025 data achieved an AUC of 0.842 on a held-out 2026
test set. A gradient boosting comparison model achieved a nearly identical AUC of 0.840,
indicating the relationship between early-game state and match outcome is close to linear.

Coefficient stability was checked across training years to assess sensitivity to patch
changes. Gold and experience differentials remained stable predictors, while
first-objective indicators shifted in importance between years; first blood specifically
lost statistical significance by 2025 and showed no predictive value in either model.
Gold differential at 15 minutes was the dominant predictor, accounting for approximately
85 percent of the gradient boosting model's total contribution.

# Part 2: Meta-Trend Forecasting

## Building the weekly series

```{r}
library(lubridate)
library(tsibble)
library(zoo)

# weekly mean |gold diff at 15 min| as a proxy for how decisive games are early on.
weekly_series <- lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  group_by(week) |>
  summarise(
    mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE),
    n_games = n()
  ) |>
  arrange(week)

range(weekly_series$week)
nrow(weekly_series)
summary(weekly_series$n_games)
```

```{r}
# sparse weeks (n_games < 10) get their mean treated as unreliable
weekly_ts <- weekly_series |>
  mutate(mean_abs_golddiff15 = if_else(n_games < 10, NA_real_, mean_abs_golddiff15)) |>
  as_tsibble(index = week) |>
  fill_gaps()

sum(is.na(weekly_ts$n_games))

weekly_ts <- weekly_ts |>
  mutate(mean_abs_golddiff15 = na.approx(mean_abs_golddiff15, na.rm = FALSE))

sum(is.na(weekly_ts$mean_abs_golddiff15))
```

```{r}
# the one remaining NA is the current, still-in-progress calendar week
weekly_ts |> filter(is.na(mean_abs_golddiff15))
```

```{r}
weekly_ts <- weekly_ts |> filter(!is.na(mean_abs_golddiff15))

sum(is.na(weekly_ts$mean_abs_golddiff15))
nrow(weekly_ts)
```

```{r}
library(ggplot2)

weekly_ts |>
  ggplot(aes(x = week, y = mean_abs_golddiff15)) +
  geom_line() +
  geom_point(data = weekly_ts |> filter(n_games < 10 | is.na(n_games)), color = "red") +
  labs(title = "Weekly mean |gold diff at 15 min|", x = NULL, y = "Gold difference")
```

## Investigating the extreme weeks

```{r}
weekly_ts |> as_tibble() |> arrange(desc(mean_abs_golddiff15)) |> slice(1:3)
weekly_ts |> as_tibble() |> arrange(mean_abs_golddiff15) |> slice(1:3)
```

```{r}
# lowest week traces to MSI. International events produce closer, less decisive early games than domestic play
low_week_start <- as.Date(min(weekly_ts$week[weekly_ts$mean_abs_golddiff15 == min(weekly_ts$mean_abs_golddiff15)]))

lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue",
         date >= low_week_start, date < low_week_start + days(7)) |>
  count(league, patch, sort = TRUE)
```

```{r}
# the highest weeks form a 3-week cluster (2024 W40-W42) 
lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  filter(week %in% yearweek(c("2024 W40", "2024 W41", "2024 W42"))) |>
  group_by(week, league) |>
  summarise(mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE), n = n(), .groups = "drop") |>
  arrange(week, desc(n))

lol_data |>
  filter(position == "team", datacompleteness == "complete", side == "Blue") |>
  mutate(week = yearweek(date)) |>
  filter(week %in% yearweek(c("2025 W27", "2025 W42", "2025 W47"))) |>
  group_by(week, league) |>
  summarise(mean_abs_golddiff15 = mean(abs(golddiffat15), na.rm = TRUE), n = n(), .groups = "drop") |>
  arrange(week, desc(n))
```

**Result:** high weeks are driven by smaller/regional leagues (bigger skill gaps -> more
early blowouts), while low weeks trace to MSI/Worlds (evenly matched elite competition
-> closer games). This pools every league in the dataset together, which is a genuine
confound worth documenting: a "meta trend" built from this mix reflects both how the top
of the competitive scene evolves and which smaller leagues happen to be scheduled that
week. Kept as-is and documented as a limitation rather than narrowing to major leagues
only, to avoid over-narrowing the analysis.

## Classical forecasting models

```{r}
library(fable)
library(feasts)
```

```{r}
weekly_ts |>
  ACF(mean_abs_golddiff15, lag_max = 20) |>
  autoplot()

weekly_ts |>
  gg_tsdisplay(mean_abs_golddiff15, plot_type = "partial")
```

ACF tails off gradually while PACF cuts off sharply after lag 1, signifying AR model signature, not
MA. KPSS test below checks stationarity.

```{r}
weekly_ts |> features(mean_abs_golddiff15, unitroot_kpss)
```

KPSS did not reject stationarity, supporting d = 0.

```{r}
fit <- weekly_ts |>
  model(
    arima_auto = ARIMA(mean_abs_golddiff15),
    arima_manual = ARIMA(mean_abs_golddiff15 ~ pdq(1, 0, 0)),
    ets = ETS(mean_abs_golddiff15)
  )

fit |> select(arima_auto) |> report()
```

```{r}
fit |> select(arima_manual) |> report()
```

Auto-selection landed on ARIMA(3,0,1); a manually specified AR(1) had a comparable AICc
(1836.76 vs. 1834.57), close enough that both looked defensible on in-sample fit alone.
The rolling-origin backtest below compares it on out-of-sample forecasting accuracy.

```{r}
weekly_cv <- weekly_ts |>
  stretch_tsibble(.init = 78, .step = 4)

cv_forecasts <- weekly_cv |>
  model(
    arima_auto   = ARIMA(mean_abs_golddiff15),
    arima_manual = ARIMA(mean_abs_golddiff15 ~ pdq(1, 0, 0)),
    ets          = ETS(mean_abs_golddiff15)
  ) |>
  forecast(h = 4)

cv_forecasts |> accuracy(weekly_ts)
```

AR(1) performed meaningfully worse out-of-sample (RMSE 327 vs. 312 for arima_auto)
despite its close in-sample AIC. ARIMA(3,0,1) and ETS ended up nearly tied as the two best models.

## Machine learning comparison

```{r}
weekly_features <- weekly_ts |>
  as_tibble() |>
  arrange(week) |>
  mutate(
    lag1 = lag(mean_abs_golddiff15, 1),
    lag2 = lag(mean_abs_golddiff15, 2),
    lag3 = lag(mean_abs_golddiff15, 3),
    lag4 = lag(mean_abs_golddiff15, 4)
  )

# first 4 rows have no history to lag from, drop them
sum(is.na(weekly_features$lag4))

weekly_features <- weekly_features |> filter(!is.na(lag4))
nrow(weekly_features)
```

```{r}
# matches the same rolling-origin folds (.init = 78, .step = 4) used for ARIMA/ETS
origins <- seq(78, nrow(weekly_features) - 4, by = 4)

fold_results <- map_dfr(origins, function(origin) {
  train_fold <- weekly_features |> slice(1:origin)
  test_fold  <- weekly_features |> slice((origin + 1):(origin + 4))

  dtrain_fold <- xgb.DMatrix(
    data = as.matrix(train_fold |> select(lag1, lag2, lag3, lag4)),
    label = train_fold$mean_abs_golddiff15
  )
  dtest_fold <- xgb.DMatrix(
    data = as.matrix(test_fold |> select(lag1, lag2, lag3, lag4))
  )

  xgb_fold <- xgb.train(
    params = xgb.params(objective = "reg:squarederror"),
    data = dtrain_fold, nrounds = 50, verbose = 0
  )

  tibble(
    week = test_fold$week,
    actual = test_fold$mean_abs_golddiff15,
    pred = predict(xgb_fold, dtest_fold)
  )
})

fold_results |>
  summarise(
    RMSE = sqrt(mean((actual - pred)^2)),
    MAE  = mean(abs(actual - pred))
  )
```

```{r}
# checking whether xgboost's RMSE advantage traces to the extreme weeks already identified above
arima_errors <- cv_forecasts |>
  filter(.model == "arima_auto") |>
  as_tibble() |>
  left_join(as_tibble(weekly_ts) |> select(week, actual = mean_abs_golddiff15), by = "week") |>
  mutate(arima_error = abs(actual - .mean)) |>
  select(week, arima_error)

xgb_errors <- fold_results |>
  mutate(xgb_error = abs(actual - pred)) |>
  select(week, xgb_error)

error_compare <- arima_errors |>
  inner_join(xgb_errors, by = "week") |>
  mutate(xgb_advantage = arima_error - xgb_error) |>
  arrange(desc(xgb_advantage))

error_compare |> print(n = 15)
```

**Summary:** Weekly mean early-game gold differential (|golddiffat15|) was constructed as
a proxy for how decisive professional games are by the 15-minute mark, then forecast
using classical and machine learning approaches with rolling-origin backtesting. Extreme
weeks were investigated directly to identify their underlying causes. Unusually high
weeks were driven by smaller regional leagues, while unusually low weeks corresponded to
international events (MSI, Worlds), where closely matched competition produces less
decisive early games. This league-composition effect is documented as a known limitation
of the pooled series.

Model order was identified via ACF/PACF inspection and confirmed against automatic ARIMA
selection; a manually specified AR(1) model had comparable in-sample AIC to the selected
ARIMA(3,0,1) but performed meaningfully worse on out-of-sample backtesting (RMSE 327 vs.
312), illustrating that in-sample fit criteria do not reliably predict forecasting
accuracy. ARIMA(3,0,1) and an ETS model performed comparably as the strongest classical
approaches.

A gradient boosting model using lagged features was subsequently compared against both
using matched backtesting folds, achieving a notably lower RMSE (274.7 vs. 311.7), with
MAE remaining comparable. This advantage was traced to specific weeks surrounding major
international tournaments and season transitions, periods during which the series
behaves least like a smooth continuous process, consistent with a nonlinear model more
effectively capturing sharp deviations from trend that a linear autoregressive structure
underestimates.

# Part 3: Cross-Region Team Rating

## League structure and region mapping

```{r}
lol_data |>
  filter(position == "team") |>
  count(league, sort = TRUE) |>
  print(n = 30)
```

```{r}
# LCS/LTA N and LLA+CBLOL/LTA S share substantial team rosters
lcs_teams  <- lol_data |> filter(league == "LCS", position == "team") |> distinct(teamname) |> pull()
ltan_teams <- lol_data |> filter(league == "LTA N", position == "team") |> distinct(teamname) |> pull()
intersect(lcs_teams, ltan_teams)

lla_cblol_teams <- lol_data |> filter(league %in% c("LLA", "CBLOL"), position == "team") |> distinct(teamname) |> pull()
ltas_teams      <- lol_data |> filter(league == "LTA S", position == "team") |> distinct(teamname) |> pull()
intersect(lla_cblol_teams, ltas_teams)
```

```{r}
region_map <- tibble(
  league = c("LPL", "LCK", "LEC", "LCS", "LTA N", "LLA", "CBLOL", "LTA S"),
  region = c("China", "Korea", "Europe", "North America", "North America",
             "Latin America", "Latin America", "Latin America")
)

# a team's home region is whichever major league it appears in most
team_home_region <- lol_data |>
  filter(position == "team", league %in% region_map$league) |>
  left_join(region_map, by = "league") |>
  count(teamname, region, sort = TRUE) |>
  group_by(teamname) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup()

team_home_region |> count(region, sort = TRUE)
nrow(team_home_region)
```

```{r}
# one row per game, chronologically ordered 
game_results <- lol_data |>
  filter(position == "team") |>
  select(gameid, date, side, teamname, result) |>
  pivot_wider(names_from = side, values_from = c(teamname, result)) |>
  mutate(winner = if_else(result_Blue == 1, teamname_Blue, teamname_Red)) |>
  select(gameid, date, team_blue = teamname_Blue, team_red = teamname_Red, winner) |>
  arrange(date)

nrow(game_results)
sum(is.na(game_results$date))
```

## Elo rating modeling

```{r}
# K-factor tested empirically (one-step-ahead Brier score)
run_elo <- function(K) {
  teams <- unique(c(game_results$team_blue, game_results$team_red))
  ratings <- setNames(rep(1500, length(teams)), teams)
  n_games <- nrow(game_results)
  e_blue_vec <- numeric(n_games)
  actual_blue_win <- numeric(n_games)

  for (i in seq_len(n_games)) {
    blue <- game_results$team_blue[i]
    red  <- game_results$team_red[i]
    r_blue <- ratings[[blue]]
    r_red  <- ratings[[red]]

    e_blue <- 1 / (1 + 10^((r_red - r_blue) / 400))
    s_blue <- if (game_results$winner[i] == blue) 1 else 0

    e_blue_vec[i] <- e_blue
    actual_blue_win[i] <- s_blue

    ratings[[blue]] <- r_blue + K * (s_blue - e_blue)
    ratings[[red]]  <- r_red + K * ((1 - s_blue) - (1 - e_blue))
  }

  mean((e_blue_vec - actual_blue_win)^2)
}

results_by_K <- map_dfr(c(8, 16, 24, 32, 40, 48, 56), function(k) {
  tibble(K = k, brier_score = run_elo(k))
})

results_by_K
```

```{r}
# K=24 vs K=48 (best Brier score): top tier is identical, but mid-table rankings reshuffle  
run_elo_full <- function(K) {
  teams <- unique(c(game_results$team_blue, game_results$team_red))
  ratings <- setNames(rep(1500, length(teams)), teams)
  n_games <- nrow(game_results)

  for (i in seq_len(n_games)) {
    blue <- game_results$team_blue[i]
    red  <- game_results$team_red[i]
    r_blue <- ratings[[blue]]
    r_red  <- ratings[[red]]
    e_blue <- 1 / (1 + 10^((r_red - r_blue) / 400))
    s_blue <- if (game_results$winner[i] == blue) 1 else 0
    ratings[[blue]] <- r_blue + K * (s_blue - e_blue)
    ratings[[red]]  <- r_red + K * ((1 - s_blue) - (1 - e_blue))
  }

  tibble(teamname = names(ratings), rating = unname(ratings))
}

ratings_24 <- run_elo_full(24) |> rename(rating_24 = rating)
ratings_48 <- run_elo_full(48) |> rename(rating_48 = rating)

compare_ratings <- ratings_24 |> left_join(ratings_48, by = "teamname")

cor(compare_ratings$rating_24, compare_ratings$rating_48, method = "spearman")

compare_ratings |> arrange(desc(rating_24)) |> slice(1:10) |> select(teamname, rating_24)
compare_ratings |> arrange(desc(rating_48)) |> slice(1:10) |> select(teamname, rating_48)
```

```{r}
# checked whether T1's rank change under higher K traces to a real recent form dip
game_results |>
  filter(team_blue == "T1" | team_red == "T1") |>
  mutate(t1_won = (team_blue == "T1" & winner == "T1") | (team_red == "T1" & winner == "T1")) |>
  arrange(date) |>
  mutate(game_num = row_number()) |>
  group_by(period = ntile(game_num, 4)) |>
  summarise(win_rate = mean(t1_won), n = n())
```

K = 24 retained: the conventional value, empirically close to the Brier-optimal K=48, and the top competitive tier is identical under both.

## Regional comparison

```{r}
region_ratings <- team_home_region |>
  left_join(ratings_24, by = "teamname")

region_ratings |>
  group_by(region) |>
  summarise(
    mean_rating = mean(rating_24),
    median_rating = median(rating_24),
    max_rating = max(rating_24),
    n_teams = n()
  ) |>
  arrange(desc(mean_rating))

region_ratings |>
  ggplot(aes(x = reorder(region, rating_24, median), y = rating_24)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "Team Elo ratings by region", x = NULL, y = "Elo rating")
```

```{r}
# top-3 average as a cleaner proxy for elite-tier strength than a mean that a few weak teams can offset
region_ratings |>
  group_by(region) |>
  slice_max(rating_24, n = 3) |>
  summarise(top3_mean = mean(rating_24), .groups = "drop") |>
  arrange(desc(top3_mean))

set.seed(1)
bootstrap_region_means <- map_dfr(1:2000, function(i) {
  region_ratings |>
    group_by(region) |>
    slice_sample(prop = 1, replace = TRUE) |>
    summarise(mean_rating = mean(rating_24), .groups = "drop")
})

bootstrap_region_means |>
  group_by(region) |>
  summarise(lower_95 = quantile(mean_rating, 0.025), upper_95 = quantile(mean_rating, 0.975)) |>
  arrange(desc(lower_95))
```

**Summary:** Team strength was estimated using a sequentially updated Elo rating system
spanning all professional matches, including domestic league play and international
events (MSI, Worlds), which serve as the primary bridge for comparing teams across
regions that otherwise rarely compete against one another. Prior to model construction,
a structural change in the Americas competitive scene was identified and confirmed
through team-roster overlap: LCS, LLA, and CBLOL were consolidated into a unified LTA
system mid-dataset, requiring these leagues to be treated as continuous regions.

The K-factor governing rating sensitivity was evaluated empirically across a range of
values using one-step-ahead prediction accuracy (Brier score) instead of a default
assumption. Results indicated the top competitive tier is stable regardless of K-factor
choice, while mid-table rankings show meaningful sensitivity to how heavily recent
results are weighted. K=24, the conventional value, was retained given its comparable
predictive performance and standard interpretability.

Regional comparison was conducted using bootstrapped confidence intervals instead of
point estimates alone. Raw mean ratings suggested a clear regional hierarchy led by
Korea; however, confidence intervals for all five regions overlapped substantially,
indicating this apparent ranking is not statistically well-supported given the limited
number of teams per region. A top-3 average provided a more defensible measure of
elite-tier strength, indicating Korea's advantage over other regions is concentrated in
genuine depth at the top. China's high maximum rating, by contrast, reflects a single
standout team instead of broad regional strength.

# Part 4: Patch-Impact Analysis

## Patch identifier integrity

```{r}
# patches use zero-padded notation (e.g. 14.01, 14.10) 
raw_patches <- map_dfr(list.files("data/raw", full.names = TRUE), function(f) {
  read_csv(f, col_types = cols_only(gameid = col_character(), patch = col_character()))
})

raw_patches |> distinct(patch) |> arrange(patch) |> print(n = Inf)
```

```{r}
# a small number of games (12 of 26,556) have no patch recorded. They are excluded.
raw_patches |>
  distinct(gameid, .keep_all = TRUE) |>
  summarise(n_missing = sum(is.na(patch)), n_total = n(), pct_missing = mean(is.na(patch)))
```

## Locating the shift

```{r}
patch_herald_tower <- lol_data |>
  filter(position == "team", side == "Blue", !is.na(patch)) |>
  group_by(patch) |>
  summarise(
    win_rate_given_herald = mean(result[firstherald == 1], na.rm = TRUE),
    win_rate_given_tower  = mean(result[firsttower == 1], na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(patch)

summary(patch_herald_tower$n_games)
patch_herald_tower |> filter(n_games < 30) |> print(n = Inf)
```

```{r}
# excluding sparse patches (n < 30) and using an ordered factor axis
patch_herald_tower_clean <- patch_herald_tower |>
  filter(n_games >= 30) |>
  arrange(patch) |>
  mutate(patch_factor = factor(patch, levels = unique(patch), ordered = TRUE))

patch_herald_tower_clean |>
  pivot_longer(cols = c(win_rate_given_herald, win_rate_given_tower), names_to = "objective", values_to = "win_rate") |>
  ggplot(aes(x = patch_factor, y = win_rate, color = objective, group = objective)) +
  geom_line() +
  geom_point() +
  labs(title = "Win rate given first herald / first tower, by patch (n \u2265 30 games)", x = "Patch", y = "Win rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r}
win_prediction_patched <- win_prediction_data |>
  mutate(season = floor(patch))

season_coefs <- map_dfr(c(14, 15, 16), function(s) {
  m <- glm(
    result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald,
    data = win_prediction_patched |> filter(season == s), family = binomial
  )
  tibble(season = s, term = names(coef(m)), estimate = coef(m))
})

season_coefs |> filter(term %in% c("firsttower", "firstherald")) |> arrange(term, season)
```

Both objectives roughly double in weight from season 14 to 15, confirming the original
finding wasn't a calendar-year aspect. Season 16 diverges: firsttower's elevated
importance persists, while firstherald partially reverts.

## Formal testing

```{r}
# likelihood-ratio test for an objective-by-period interaction 
boundary_data <- win_prediction_patched |>
  filter(season %in% c(14, 15)) |>
  mutate(period = factor(season, levels = c(14, 15), labels = c("before", "after")))

m_null <- glm(
  result ~ golddiffat15 + xpdiffat15 + csdiffat15 + firstblood + firstdragon + firsttower + firstherald + period,
  data = boundary_data, family = binomial
)

m_tower_int  <- update(m_null, . ~ . + firsttower:period)
m_herald_int <- update(m_null, . ~ . + firstherald:period)

tower_test  <- anova(m_null, m_tower_int, test = "Chisq")
herald_test <- anova(m_null, m_herald_int, test = "Chisq")

tower_p  <- tower_test$`Pr(>Chi)`[2]
herald_p <- herald_test$`Pr(>Chi)`[2]

# FDR correction since both objectives are being tested
p.adjust(c(firsttower = tower_p, firstherald = herald_p), method = "fdr")
```

**Summary:** This analysis investigated an anomaly identified in Part 1: the predictive
weight of first herald and first tower on match outcome nearly doubled between two
training years, while first blood's effect collapsed to non-significance. Before
conducting patch-level analysis, the integrity of the patch identifier was verified
directly, confirming that zero-padded patch notation (e.g., 14.01, 14.10) avoids a
numeric collision risk that would otherwise be introduced by floating-point storage of
version numbers.

Season-level coefficients from Part 1's original model were recomputed using
patch-derived season boundaries instead of calendar years. Both objectives showed a
comparable increase in predictive weight between seasons 14 and 15, confirming the
original finding was not an artifact of the calendar-year split. Season 16 revealed a
divergence: first tower's elevated importance persisted, while first herald's effect
partially reverted toward its original level, indicating two distinct underlying
dynamics rather than a single shared cause.

The season 14-to-15 shift was formally tested using likelihood-ratio tests for
objective-by-period interaction terms within the original logistic regression, the
appropriate test given that the original finding concerned a controlled coefficient
rather than a marginal win rate. Following FDR correction across both objectives, both
interactions were highly significant (first tower: p = 2.47 x 10^-13; first herald:
p = 1.11 x 10^-9), confirming the coefficient shifts identified in Part 1 reflect genuine
changes in game dynamics rather than sampling variation.
